Cooling Requirements for the Vertical Shear Instability in Protoplanetary Disks

### 2015-05-08 paper.tex
minor

For a thin disk ($h\ll 1$), the vertical shear rate varies with height, $z$, from the midplane as 
\begin{align}\label{vshear_thin}
  r \frac{\p \Omega }{\p z} \simeq  h q \OmK \left(\frac{z}{2 H}\right),
\end{align}
where $H=hr$ is the characteristic disk scale height. 
%for several disk scale heights, $H = hr$. 
%As noted in
%\S\ref{intro} this vertical variation is a source of free energy and
%thus vertical shear ($\p_z\Omega \neq0$) is destabilizing. 
%
%For $\gamma>1$ the disk is stably stratified with a real 
This destabilizing shear competes with the stabilizing vertical
buoyancy frequency, $N_z$.  In a thin disk,     
\begin{align}\label{nz_thin}
  N_z^2 \simeq \left(\frac{\gamma-1}{\gamma}\right) \left(\frac{z}{H}\right)^2
  \OmK^2.  
\end{align}
Fig. \ref{eqm_structure} maps the vertical shear rate and buoyancy
frequency (as well as the gas density) 
in a fiducial PPD model, see \S\ref{toy_relax} for details. 
%, whose parameters are in Appendix \ref{mmsn} . %Here $\rho_0$
%is the midplane density and vertical shear rate and buoyancy
%frequencies are normalized by the local Keplerian frequency. 
Vertical shear is 
generally weak compared to buoyancy, suggesting stability. In fact, 
without any cooling, the Solberg-Hoiland criteria \citep{tassoul78}  
confirms that vertical buoyancy is strong
enough to ensure (axisymmetric) stability, see \S\ref{solberg}.  With radiative cooling, 
thermal fluctuations decay which, combined with pressure equilibration, reduces the effective buoyancy.

%However, the stabilizing effect of buoyancy can be mitigated for 
%effectively isothermal disturbances. Consider a fluid element displaced upwards, away from
%the disk midplane. In a stably stratified adiabatic disk, it would
%sink because of buoyancy. However, if thermal timescales are short, in
%that temperature flucations decay rapidly, then the fluid element would attain its surrounding 
%density and  
%%(assuming an ideal gas and it reaches pressure 
%%equilibrium with its surroundings). 
%thus buoyancy is rendered ineffective. Qualitatively, a short thermal
%timescale $t_c$ reduces the effective (stabilizing) buoyancy force.

How short must $t_c$ be for vertical shear to prevail?  We start with 
 the Richardson number $\mathrm{Ri} = N_z^2/(r\p_z\Omega)^2$, a 
ratio of buoyant to shear energies.  Though not precisely applicable to our problem,
non-rotating shear flows are stable if $\mathrm{Ri} > 1/4$ (\citealp{chandrasekhar61}, 
also see \citealp{ys02, lee10} for applications to thin dust layers in PPDs).
Following \cite{urpin03} and \cite{townsend58},  we reduce the buoyant energy by the 
ratio of cooling to forcing timescales, $t_c |r \p_z\Omega| \leq 1$ (with no reduction in buoyant energy 
expected when this inequality fails).  With this reduction, the Richardson-like criterion for 
shear instability becomes  \citep{urpin03}
\begin{align}\label{bcrit_est}
  t_c \lesssim \frac{\left|r\p\Omega/\p z\right|} {N_z^2}.
\end{align}
 %This result is far from rigorous, but it contains some useful insights.
If we interpret Eq.\ \ref{bcrit_est} as a local criterion, then we see that 
the maximum cooling time which permits instability decreases with height as $1/|z|$;  
the stabilizing effect of vertical buoyancy increases more rapidly away from
the midplane than the destabilizing effect of vertical shear.  
This finding is relevant to the radiative damping of so called `surface 
modes', which we consider in \S\ref{numerical}.

We are more interested, however, the the global instability criterion for disturbances at all heights.
Our desired result 
is the cooling requirement 
\begin{align}\label{prelim_bcrit}
  t_c <  &\underbrace{\frac{h |q| }{\gamma - 1}}_{\beta_\mathrm{crit}}\OmK^{-1}
 % &\equiv \beta_\mathrm{crit}\OmK^{-1},\notag
\end{align}   
for VSI growth that we derive in \S\ref{iso_vsi_beta_crit}.  We can simply 
(but non-rigourously) obtain Eq. \ref{prelim_bcrit}
  by evaulating  Eq. \ref{bcrit_est} at $|z|=\gamma H/2$. %i.e., at $z=O(H)$

%This estimate agrees with 
%our central result, the cooling criterion that we derive more rigorously in \S\ref{iso_vsi_beta_crit}.

%
%The left-hand side represents an effective buoyancy, which is
%stabilizing; while the right-hand side represents vertical shear,
%which is destabilizing. In a thin disk $ h \ll 1$, so the destabilizing effect of 
%vertical shear is rather weak because $|r\p_z\Omega|\sim h\OmK
%\ll \OmK$. By comparison, the buoyancy frequency $N_z\sim
%\OmK$. It follows that $t_c\lesssim O( h\OmK^{-1})$ is
%required for instability.  \emph{In a thin  
%  disk the thermal timescale must be shorter than the dynamical
%  timescale by $O( h)\ll 1$ for the VSI to operate.}   
%
%
%Then even a small (positive)
%vertical entropy gradient can significantly stabilize the VSI. 
%Note that in astrophysical disks both $N_z^2$ and $\p_z\Omega$ varies
%with height. Eq. \ref{bcrit_est} then suggest that with
%increasing height, buoyancy would eventually stabilize the VSI (since
%$N_z^2\propto z^2$ and $\p_z\Omega\propto z$).
%% This means that 
%% disturbances localized away from the disk midplane are easily stabilized by
%% buoyancy. 
%This argument assumes vertically localized disturbances. It is,
%however, not clear how to apply Eq. \ref{bcrit_est} to vertically
%global disturbances, as considered in this work.
%
%
%
%
%Our central result is the requirement 
%\begin{align}\label{prelim_bcrit}
%  (\gamma - 1)t_c <  h |q| \OmK^{-1}
%\end{align}   
%for the VSI in a vertically isothermal, thin disk. This equation
%reflects Eq. \ref{bcrit_est} but there is no $z$ dependence. 
%% While assumptions and simplifications will be made 
%% to derive this, 
%Eq. \ref{prelim_bcrit} can be 
%readily applied to evaluate where in a disk the VSI is likely to
%operate.

@@

% Since cooling by radiative diffusion depends on a perturbation's lengthscale, we
% plot $t_c$ for two values of the radial wavenumber,  $k_x$,  the relevant 
% component for VSI modes that are radially narrow.  The red line
%marks the boundary implied by Eq. \ref{prelim_bcrit}.  We caution that 
% Eq. \ref{prelim_bcrit} applies when $t_c$ is vertically uniform.  This section is 
% aimed at basic insights; while more realistic spatial variations in
% $t_c$ are considered in \S\ref{application}. 
% %analyzes the VSI with realistic spatial variations in.
% 
% Due to the smaller values of $t_c$ in the outer disk, we generally expect the VSI to 
% operate outside $\sim 10$ AU.  This conclusion changes however, if
% the dust opacity is reduced,  
%slowing down cooling, as we show in \S\ref{application}. % (e.g.\ Fig.\
%% \ref{mmsn_bcrit_bcool}). 
%In the inner disk, thermal times are long in the midplane, but short at higher altitutdes. 
%Again, this structure necessitates the more detailed analyses that follow.  Nevertheless, we can already 
%see that due to radiative diffusion higher wavenumbers can cool more rapidly.  Thus smaller scale disturbances 
%may trigger VSI at smaller disk radii. 
%
%On the other hand, one cannot obtain arbitrarily short cooling times
%by invoking arbitrarily small scales (cf. \citealt{urpin03}, 
%\citetalias{barker15}), because very small wavelengths are in the
%optically thin cooling regime,  where cooling is independent of
%lengthscale. Furthermore, 
%%while we do not include viscosity in our
%%idealized disk models, in reality 
%small-scale perturbations will be 
%prone to viscous damping, including the effective viscosity from turbulent motions.
%While our linear, inviscid calculations do not address this issue directly, we expect that 
%shorter wavelength modes will drive weaker non-linear (and non-axisymmetric) transport. 
%
%% However, we cannot and do not find that VSI will always be relevant for
%% sufficiently small perturbations because \begin{inparaenum}[(i)] \item very small wavelengths are in the optically 
%% thin cooling regime,  where cooling is independent of lengthscale;
%% and \item small scale perturbations 
%% are more prone to viscous damping, and should drive weaker transport.
%% \end{inparaenum} 
%
%We now proceed with our detailed analysis of the VSI to arrive
%at Eq. \ref{prelim_bcrit} and verify it numerically. The interested 
%reader may jump to \S\ref{application} for the application of our
%results to PPDs.     
>>>>>>> 3c00bb4e2b076179e280f2c07728086a110764ef

>>>>

For a thin disk ($h\ll 1$), the vertical shear rate varies with height, $z$, from the midplane as 
\begin{align}\label{vshear_thin}
  r \frac{\p \Omega }{\p z} \simeq  h q \OmK \left(\frac{z}{2 H}\right),
\end{align}
where $H=hr$ is the characteristic disk scale height.

This destabilizing shear competes with the stabilizing vertical
buoyancy frequency, $N_z$.  In a thin disk,     
\begin{align}\label{nz_thin}
  N_z^2 \simeq \left(\frac{\gamma-1}{\gamma}\right) \left(\frac{z}{H}\right)^2
  \OmK^2.  
\end{align}
Fig. \ref{eqm_structure} maps the vertical shear rate and buoyancy
frequency (as well as the gas density) 
in a fiducial PPD model, see \S\ref{toy_relax} for details.

Vertical shear is 
generally weak compared to buoyancy, suggesting stability. In fact, 
without any cooling, the Solberg-Hoiland criteria %\citep{tassoul78}  Tassoul cited in S-H section is enough
confirms that vertical buoyancy is strong
enough to ensure (axisymmetric) stability, see \S\ref{solberg}.  With radiative cooling, 
thermal fluctuations decay which, combined with pressure equilibration, reduces the effective buoyancy.

How short must $t_c$ be for vertical shear to prevail?  We start with 
 the Richardson number $\mathrm{Ri} = N_z^2/(r\p_z\Omega)^2$, a 
ratio of buoyant to shear energies.  Though not precisely applicable to our problem,
non-rotating shear flows are stable if $\mathrm{Ri} > 1/4$ (\citealp{chandrasekhar61}, 
also see \citealp{ys02, lee10} for applications to thin dust layers in PPDs).
Following \cite{urpin03} and \cite{townsend58},  we reduce the buoyant energy by the 
ratio of cooling to forcing timescales, $t_c |r \p_z\Omega| \leq 1$ (with no reduction in buoyant energy 
expected when this inequality fails).  With this reduction, the Richardson-like criterion for 
shear instability becomes  \citep{urpin03}
\begin{align}\label{bcrit_est}
  t_c \lesssim \frac{\left|r\p\Omega/\p z\right|} {N_z^2}.
\end{align}
If we interpret Eq.\ \ref{bcrit_est} as a local criterion, then we see that 
the maximum cooling time which permits instability decreases with height as $1/|z|$;  
the stabilizing effect of vertical buoyancy increases more rapidly away from
the midplane than the destabilizing effect of vertical shear.  
This finding is relevant to the radiative damping of so called `surface 
modes', which we consider in \S\ref{numerical}.

We are more interested, however, the the global instability criterion for disturbances at all heights.
Our desired result 
is the cooling requirement 
\begin{align}\label{prelim_bcrit}
  t_c <  &\underbrace{\frac{h |q| }{\gamma - 1}}_{\beta_\mathrm{crit}}\OmK^{-1}
\end{align}   
for VSI growth that we derive in \S\ref{iso_vsi_beta_crit}.  We can simply 
(but non-rigourously) obtain Eq. \ref{prelim_bcrit}
  by evaulating  Eq. \ref{bcrit_est} at $|z|=\gamma H/2$.

@@

% Since cooling by radiative diffusion depends on a perturbation's lengthscale, we
% plot $t_c$ for two values of the radial wavenumber,  $k_x$,  the relevant 
% component for VSI modes that are radially narrow.  The red line
%marks the boundary implied by Eq. \ref{prelim_bcrit}.  We caution that 
% Eq. \ref{prelim_bcrit} applies when $t_c$ is vertically uniform.  This section is 
% aimed at basic insights; while more realistic spatial variations in
% $t_c$ are considered in \S\ref{application}. 
% %analyzes the VSI with realistic spatial variations in.
% 
% Due to the smaller values of $t_c$ in the outer disk, we generally expect the VSI to 
% operate outside $\sim 10$ AU.  This conclusion changes however, if
% the dust opacity is reduced,  
%slowing down cooling, as we show in \S\ref{application}. % (e.g.\ Fig.\
%% \ref{mmsn_bcrit_bcool}). 
%In the inner disk, thermal times are long in the midplane, but short at higher altitutdes. 
%Again, this structure necessitates the more detailed analyses that follow.  Nevertheless, we can already 
%see that due to radiative diffusion higher wavenumbers can cool more rapidly.  Thus smaller scale disturbances 
%may trigger VSI at smaller disk radii. 
%
%On the other hand, one cannot obtain arbitrarily short cooling times
%by invoking arbitrarily small scales (cf. \citealt{urpin03}, 
%\citetalias{barker15}), because very small wavelengths are in the
%optically thin cooling regime,  where cooling is independent of
%lengthscale. Furthermore, 
%%while we do not include viscosity in our
%%idealized disk models, in reality 
%small-scale perturbations will be 
%prone to viscous damping, including the effective viscosity from turbulent motions.
%While our linear, inviscid calculations do not address this issue directly, we expect that 
%shorter wavelength modes will drive weaker non-linear (and non-axisymmetric) transport. 
%
%% However, we cannot and do not find that VSI will always be relevant for
%% sufficiently small perturbations because \begin{inparaenum}[(i)] \item very small wavelengths are in the optically 
%% thin cooling regime,  where cooling is independent of lengthscale;
%% and \item small scale perturbations 
%% are more prone to viscous damping, and should drive weaker transport.
%% \end{inparaenum} 
%
%We now proceed with our detailed analysis of the VSI to arrive
%at Eq. \ref{prelim_bcrit} and verify it numerically. The interested 
%reader may jump to \S\ref{application} for the application of our
%results to PPDs.
